### 2020-11-18 pldi21/intro.tex
tighten up 1.0

\section{Introduction}\label{sec:intro}
Textual program editors are powerful and expressive user interfaces,
so it is perhaps little wonder that they remain dominant decades after the teletype.
However, textual user interfaces are not the best tool for every job.
In particular, there are many
data types for which a non-textual
user interface for expression construction may situationally be more appropriate.

As a simple example, consider a record type
classifying colors using an RGBA encoding:
\begin{lstlisting}[numbers=none]
type Color = { r: Int, g: Int, b: Int, a: Int }
\end{lstlisting}
It is possible to select a particular color by entering
an expression of this type using a text editor:
\begin{lstlisting}[numbers=none]
let bgcolor : Color = { r: 255, g: 178, b: 45, a: 100 }
\end{lstlisting}
The problem is that this user interface for color selection
offers the user no immediate feedback
and limited editing affordances.
It is difficult for the programmer, or anyone subsequently reading or modifying the code, to know which color is represented
and to interactively tweak that color.

Better support for manipulating data of types like these would be particularly helpful for users engaging in
live and exploratory programming in domains like web design, media production,
and data analysis. Indeed, practitioners in domains like these commonly eschew general-purpose programming
in favor of graphical end-user applications, like %
image and video editors and music composition software,
in part because these applications provide domain-specific forms of live feedback and
direct manipulation affordances, e.g. color palettes, visual timelines, and interactive plots.
The trade-off is that these applications have limited support for abstraction and composition.
It is difficult, for example, to bind a
color to a variable for use in multiple locations,
or to transform a sound by passing it through a series of functions,
or to define a function that computes a vector graphic given a set
of parameters (some of them perhaps themselves computed),
or to organize these constructions into modules.
Moreover, it is difficult to add new affordances or to compose
affordances in ways that the application designer did not anticipate,
so users cannot easily make even simple changes like replacing a numeric text box in a color chooser with a slider,
much less more ambitious changes like a new visual interface for expressing geospatial data queries.

This paper aims to resolve the apparent tension between
programmatic and direct manipulation user interfaces by designing a programming system that
is able to surface GUIs for the data types for which
they are useful, while retaining full support for symbolic program manipulation
and the abstraction and composition mechanisms
available in modern general-purpose languages.

>>>>

\section{Introduction}\label{sec:intro}
Text-based program editors are flexible and expressive user interfaces,
so it is little wonder that they remain dominant decades after the teletype.
However, textual user interfaces are not the best tool for every programming job.
% In particular, there are countless 
% data types for which a non-textual
% user interface may situationally be more appropriate.

As a simple example, consider a record type
classifying colors using the RGBA encoding. 
It is possible to select a particular color by entering
an expression of this type using a text editor, e.g. \li{\{ r: 255, g: 178, b: 45, a: 100 \}}. 
The problem with this textual user interface for color selection is that 
it offers the programmer no live feedback about which color has been selected 
and limited editing affordances for tweaking the chosen color.
Analagous critiques apply to strictly textual user interfaces for 
countless other data structures,
such as animation parameters,
vector graphics,
musical sequences,
audio filters,
game states, 
GUI primitives,
tabular data, 
plots,
geospatial data, 
neural network diagrams, 
biological neuron models, 
and the list goes on.

% It is difficult for the programmer, or anyone subsequently reading or modifying the code, to know which color is represented
% and to interactively tweak that color.

Practitioners in domains where manipulating data of types like these is central 
have largely eschewed general-purpose programming languages 
in favor of more specialized graphical end-user applications, like %
image and video editors, music composition software, level design tools, 
and bespoke GUIs written by technicians, 
in large part because these applications 
take seriously the need for domain-specific forms of live feedback, graphical representations, and
direct manipulation affordances, 
e.g. color palettes, visual timelines, interactive plots, and maps.

The tragic trade-off is that these applications have limited support for abstraction and composition.
It is difficult, for example, to bind a
color to a variable for use in multiple locations,
or to define a function that computes a vector graphic given a set
of parameters (some of them perhaps themselves computed),
or to transform a sound by passing it through a series of functions,
or to organize these constructions into modules.
Moreover, it is difficult to add new affordances or to compose
affordances in ways that the application developer did not anticipate,
so users cannot easily make even simple changes like replacing a numeric text box in a color chooser with a slider,
much less more ambitious changes like a new visual interface for expressing geospatial data queries.

% Better support for manipulating data of types like these would be particularly helpful for users engaging in
% live and exploratory programming in domains like web design, media production,
% and data analysis. Indeed,

This paper aims to resolve the apparent tension between
programmatic and direct manipulation user interfaces by designing a programming system that
is able to surface GUIs for the data types for which
they are useful, while retaining full support for symbolic program manipulation
and the abstraction and composition mechanisms
available in modern general-purpose languages as the ``glue'' between these 
GUIs.
